Gaze-dependent evidence accumulation predicts multi-alternative risky choice behaviour

### 2021-02-12 manuscript/manuscript_refs-apa.tex
revised author contributions, acknowledgments, affiliations, added conflict of interest statement, renamed files

\author[1,2,3,*]{\small Felix Molter}
\author[2,4]{\small Armin W. Thomas}
\author[5,6]{\small Scott A. Huettel}
\author[2]{\small Hauke R. Heekeren}
\author[1,2,3]{\small Peter N. C. Mohr}
\affil[1]{\footnotesize School of Business \& Economics, Freie Universität Berlin, Berlin, Germany}
\affil[2]{\footnotesize Center for Cognitive Neuroscience, Freie Universität Berlin, Berlin, Germany}
\affil[3]{\footnotesize WZB Berlin Social Science Center, Berlin, Germany}
\affil[4]{\footnotesize Department of Electrical Engineering and Computer Science, Technische Universität Berlin, Berlin, Germany}
\affil[5]{\footnotesize Center for Cognitive Neuroscience, Duke University, Durham, NC, USA}
\affil[6]{\footnotesize Department for Psychology and Neuroscience, Duke University, Durham, NC, USA}
\affil[*]{\footnotesize E-mail: \href{mailto:felixmolter@gmail.com}{felixmolter@gmail.com}}

@@

% Fig. 1: Choice task and attribute space
\begin{SCfigure}[0.8]
\caption{\textbf{Choice task and attribute space of stimuli.} \textbf{(a)} Choice task. Trials started with a central fixation cross for 1.5 s. Next, three gambles were presented. Participants made a choice by button press, without time limit. Eye movements were recorded during the choice phase. Finally, brief feedback on the choice (but not on the gamble outcome) was displayed. \textbf{(b)} Attribute space. Each gamble is described by two attributes: Probability $p$ and outcome $m$. The core options $A$ and $B$ were presented in every trial, along with one of four decoy alternatives: Compromise decoys ($C_A$ or $C_B$) and asymmetrically dominated decoys ($D_A$ or $D_B$) are expected to elicit compromise and attraction effects, respectively. Dashed lines indicate possible stimulus placements. \textbf{(c)} Exact positions for $C_B$, $A$ and $C_A$ were calibrated for pairwise indifference in separate binary choice estimation blocks for each individual. Differences in outcome between neighbouring alternatives was not less than 2 EUR. Dominated decoys $D_A$ and $D_B$ were always 2\% and 1 EUR worse than $A$ and $B$ respectively.}
\includegraphics[scale=1]{../figures/1-task_stimuli.pdf}
\hspace{0.5cm}
\label{fig:task}
\end{SCfigure}

@@

The GLA variant also described 17 of 40 (42.5\%) participants best (Table \ref{tab:switchboard-best-ind}). Another similar variant with additive attribute integration described an additional four participants best. An additive relationship between attributes is typically assumed by models of multi-attribute multi-alternative choice \parencite[e.g.,][]{roe2001MultialternativeDecisionField,usher2004LossAversionInhibition}. Furthermore, additive integration of probability and outcome has recently been suggested as an alternative to multiplicative mechanisms and has been demonstrated to be equivalent for particular parameterizations and parameter values \parencite{rouault2019PrefrontalMechanismsCombining}. Five participants were best described by variants similar to the GLA, but using gaze-dependent leakage or inhibition instead of an alternative-wise gaze-discount. Note that gaze-dependent inhibition and leakage mechanisms can act similarly to an alternative-wise gaze-discount: All three mechanisms effectively reduce accumulated evidence of non-fixated alternatives. While the alternative-wise gaze bias mechanism applies a constant discount to the accumulation inputs, the gaze-dependent inhibition mechanism is proportional to the accumulated evidence of the currently fixated alternative, and applies to the already accumulated evidence of other options, not the inputs to the accumulation process. Gaze-dependent leakage similarly reduces already accumulated evidence, proportional to the momentary accumulator value.

@@

\section*{Acknowledgements}
F.M. is supported by the International Max Planck Research School on the Life Course (LIFE). A.T. is supported by Stanford Data Science.

\section*{Author contributions}
F.M., S.A.H. and P.N.C.M. designed the study. F.M. and A.W.T. conceptualized the behavioural models. F.M. performed the experiment, wrote all software, analysed the data and visualized results. P.N.C.M. and H.R.H. supervised the project. P.N.C.M. acquired funding. F.M. wrote the original draft. All authors reviewed and edited the manuscript.

>>>>

\author[1,2,3,*]{\small Felix Molter}
\author[2,4]{\small Armin W. Thomas}
\author[5,6]{\small Scott A. Huettel}
\author[2,7]{\small Hauke R. Heekeren}
\author[1,2,3]{\small Peter N. C. Mohr}
\affil[1]{\footnotesize School of Business \& Economics, Freie Universität Berlin, Berlin, Germany}
\affil[2]{\footnotesize Center for Cognitive Neuroscience, Freie Universität Berlin, Berlin, Germany}
\affil[3]{\footnotesize WZB Berlin Social Science Center, Berlin, Germany}
\affil[4]{\footnotesize Department of Electrical Engineering and Computer Science, Technische Universität Berlin, Berlin, Germany}
\affil[5]{\footnotesize Center for Cognitive Neuroscience, Duke University, Durham, NC, USA}
\affil[6]{\footnotesize Department for Psychology and Neuroscience, Duke University, Durham, NC, USA}
\affil[7]{\footnotesize Department for Education and Psychology, Freie Universität Berlin, Berlin, Germany}
\affil[*]{\footnotesize E-mail: \href{mailto:felixmolter@gmail.com}{felixmolter@gmail.com}}

@@

% Fig. 1: Choice task and attribute space
\begin{SCfigure}[0.8][ht]
\caption{\textbf{Choice task and attribute space of stimuli.} \textbf{(a)} Choice task. Trials started with a central fixation cross for 1.5 s. Next, three gambles were presented. Participants made a choice by button press, without time limit. Eye movements were recorded during the choice phase. Finally, brief feedback on the choice (but not on the gamble outcome) was displayed. \textbf{(b)} Attribute space. Each gamble is described by two attributes: Probability $p$ and outcome $m$. The core options $A$ and $B$ were presented in every trial, along with one of four decoy alternatives: Compromise decoys ($C_A$ or $C_B$) and asymmetrically dominated decoys ($D_A$ or $D_B$) are expected to elicit compromise and attraction effects, respectively. Dashed lines indicate possible stimulus placements. \textbf{(c)} Exact positions for $C_B$, $A$ and $C_A$ were calibrated for pairwise indifference in separate binary choice estimation blocks for each individual. Differences in outcome between neighbouring alternatives was not less than 2 EUR. Dominated decoys $D_A$ and $D_B$ were always 2\% and 1 EUR worse than $A$ and $B$ respectively.}
\includegraphics[scale=1]{../figures/1-task_stimuli.pdf}
\hspace{0.5cm}
\label{fig:task}
\end{SCfigure}

@@

The GLA variant also described 17 of 40 (42.5\%) participants best (Table \ref{tab:switchboard-best-ind}). Another similar variant with additive attribute integration described an additional four participants best. An additive relationship between attributes is typically assumed by models of multi-attribute multi-alternative choice \parencite[e.g.,][]{roe2001MultialternativeDecisionField,usher2004LossAversionInhibition}. Furthermore, additive integration of probability and outcome has recently been suggested as an alternative to multiplicative mechanisms and has been demonstrated to be equivalent for particular parameterisations and parameter values \parencite{rouault2019PrefrontalMechanismsCombining}. Five participants were best described by variants similar to the GLA, but using gaze-dependent leakage or inhibition instead of an alternative-wise gaze-discount. Note that gaze-dependent inhibition and leakage mechanisms can act similarly to an alternative-wise gaze-discount: All three mechanisms effectively reduce accumulated evidence of non-fixated alternatives. While the alternative-wise gaze bias mechanism applies a constant discount to the accumulation inputs, the gaze-dependent inhibition mechanism is proportional to the accumulated evidence of the currently fixated alternative, and applies to the already accumulated evidence of other options, not the inputs to the accumulation process. Gaze-dependent leakage similarly reduces already accumulated evidence, proportional to the momentary accumulator value.

@@

\section*{Acknowledgements}
F.M. is supported by the International Max Planck Research School on the Life Course (LIFE). A.T. is supported by the Max Planck School of Cognition and Stanford Data Science.

\section*{Author contributions}
F.M., S.A.H. and P.N.C.M. designed the study. F.M. and A.W.T. conceptualized the behavioural models. F.M. perform ed the experiment, wrote all software, analysed the data and visualized results. P.N.C.M. and H.R.H. supervised the project. P.N.C.M. acquired funding. F.M. wrote the original draft. All authors reviewed and edited the manuscript.

\section*{Competing interests}
The authors declare no competing interests.

### 2021-02-12 manuscript/manuscript_refs-numbered.tex
revised author contributions, acknowledgments, affiliations, added conflict of interest statement, renamed files

% Load packages
\usepackage{booktabs}  % Nicer tables
\usepackage{rotating}  % For rotated table
\usepackage[table,xcdraw]{xcolor}  % Coloring
\usepackage{lineno}    % Line numbers
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage[babel]{csquotes}
\usepackage{amsmath}
\usepackage{textcomp}
\usepackage{gensymb}   % for \degree
\usepackage{graphicx}
\usepackage{authblk}   % Multiple authors with affiliations

@@

\author[1,2,*]{\small Felix Molter}
\author[2,3,4]{\small Armin W. Thomas}
\author[5,6]{\small Scott A. Huettel}
\author[2]{\small Hauke R. Heekeren}
\author[1,2]{\small Peter N. C. Mohr}
\affil[1]{\footnotesize School of Business \& Economics, Freie Universität Berlin, Berlin, Germany}
\affil[2]{\footnotesize Center for Cognitive Neuroscience, Freie Universität Berlin, Berlin, Germany}
\affil[3]{\footnotesize Stanford Data Science, Stanford University, Stanford, CA, USA}
\affil[4]{\footnotesize Department of Psychology, Stanford University, Stanford, CA, USA}
\affil[5]{\footnotesize Center for Cognitive Neuroscience, Duke University, Durham, NC, USA}
\affil[6]{\footnotesize Department for Psychology and Neuroscience, Duke University, Durham, NC, USA}
\affil[*]{\footnotesize E-mail: \href{mailto:felixmolter@gmail.com}{felixmolter@gmail.com}}

@@

% Fig. 1: Choice task and attribute space
\begin{figure}[t!]
\centering
\includegraphics[scale=1]{../figures/1-task_stimuli.pdf}
\caption{\textbf{Choice task and attribute space of stimuli.} \textbf{(a)} Choice task. Trials started with a central fixation cross for 1.5 s. Next, three gambles were presented. Participants made a choice by button press, without time limit. Eye movements were recorded during the choice phase. Finally, brief feedback on the choice (but not on the gamble outcome) was displayed. \textbf{(b)} Attribute space. Each gamble is described by two attributes: Probability $p$ and outcome $m$. The core options $A$ and $B$ were presented in every trial, along with one of four decoy alternatives: Compromise decoys ($C_A$ or $C_B$) and asymmetrically dominated decoys ($D_A$ or $D_B$) are expected to elicit compromise and attraction effects, respectively. Dashed lines indicate possible stimulus placements. \textbf{(c)} Exact positions for $C_B$, $A$ and $C_A$ were calibrated for pairwise indifference in separate binary choice estimation blocks for each individual. Differences in outcome between neighbouring alternatives was not less than 2 EUR. Dominated decoys $D_A$ and $D_B$ were always 2\% and 1 EUR worse than $A$ and $B$ respectively.}
\label{fig:task}
\end{figure}

@@

The GLA variant also described 17 of 40 (42.5\%) participants best (Table \ref{tab:switchboard-best-ind}). Another similar variant with additive attribute integration described an additional four participants best. An additive relationship between attributes is typically assumed by models of multi-attribute multi-alternative choice \autocite{roe2001MultialternativeDecisionField,usher2004LossAversionInhibition}. Furthermore, additive integration of probability and outcome has recently been suggested as an alternative to multiplicative mechanisms and has been demonstrated to be equivalent for particular parameterizations and parameter values \autocite{rouault2019PrefrontalMechanismsCombining}. Five participants were best described by variants similar to the GLA, but using gaze-dependent leakage or inhibition instead of an alternative-wise gaze-discount. Note that gaze-dependent inhibition and leakage mechanisms can act similarly to an alternative-wise gaze-discount: All three mechanisms effectively reduce accumulated evidence of non-fixated alternatives. While the alternative-wise gaze bias mechanism applies a constant discount to the accumulation inputs, the gaze-dependent inhibition mechanism is proportional to the accumulated evidence of the currently fixated alternative, and applies to the already accumulated evidence of other options, not the inputs to the accumulation process. Gaze-dependent leakage similarly reduces already accumulated evidence, proportional to the momentary accumulator value.

@@

% Fig. 5: Hybrid variant
\begin{figure}[t!]
\centering
\includegraphics[scale=1]{../figures/5-hybrid-variant.pdf}
\caption{\textbf{Hybrid model variant details.} \textbf{(a, b)} Number of participants better described by the hybrid variant (pink) or the GLA (grey), dependent on strength of attraction \textbf{(a)} and compromise \textbf{(b)} effects. Participants with strong attraction effects were better described by the hybrid variant. \textbf{(c, d)} Individual observed and predicted RST in attraction \textbf{(c)} and compromise trials \textbf{(d)}. Compared to GLA (Fig. \ref{fig:modelcomparison}i, j), the hybrid model better predicted strong attraction effects for some participants. Predictions of compromise effects are similar. \textbf{(e, f)} Observed and model-predicted probability of choosing the target alternative, depending on the target’s relative dwell time advantage. Like other gaze-dependent models (Fig. \ref{fig:modelcomparison}), the hybrid variant generally captured the positive association between gaze and choice. In contrast to GLA, however, it predicted an overall higher probability of choosing the target in attraction trials \textbf{(e)}. Predictions in compromise trials \textbf{(f)} are similar to GLA. White bars and error bars indicate mean $\pm$ s.e. observed data from even-numbered trials. Model predictions are based on 50 simulations for each odd-numbered trial.}
\label{fig:hybrid}
\end{figure}

@@

\section*{Acknowledgements}
F.M. is supported by the International Max Planck Research School on the Life Course (LIFE). A.T. is supported by Stanford Data Science.

\section*{Author contributions}
F.M., S.A.H. and P.N.C.M. designed the study. F.M. and A.W.T. conceptualized the behavioural models. F.M. performed the experiment, wrote all software, analysed the data and visualized results. P.N.C.M. and H.R.H. supervised the project. F.M. wrote the original draft. All authors reviewed and edited the manuscript.

@@

% Supplementary Fig.: Dwell-regression weights
\begin{figure}[t!]
\centering
\includegraphics[scale=1]{../figures/S_dwell-regression_weights.pdf}
\caption{\textbf{Weight estimates from regression analyses on absolute dwell times.} We performed two mixed-effects regression analyses of dwell time towards each AOI onto stimulus properties: \textbf{(a, c)} Regressing the total dwell time towards an AOI over a full trial onto AOI column, row, attribute rank (best, middle or worst value on the attribute), two dummy predictors coding alternative, attribute (probability or outcome) and whether the AOI belonged to the subsequently chosen alternative. \textbf{(b, d)} Second, we binned dwell times in each trial into five time-bins and added an interaction term with time-bin for each predictor. The panels show the interaction term weights. Analyses were carried out separately for attraction \textbf{(a, b)} and compromise \textbf{(c, d)} trials. Regression models had random intercepts and slopes across participants. Points and intervals mark posterior mean estimates and associated HDI$_{95}$ (coloured green if the interval excluded 0).}
\label{fig:dwell-regression-weights}
\end{figure}

@@

\begin{sidewaystable}[ht]

@@

\end{sidewaystable}

>>>>

% Load packages
\usepackage{booktabs}  % Nicer tables
\usepackage{lscape}
\usepackage[table,xcdraw]{xcolor}  % Coloring
\usepackage{lineno}    % Line numbers
\usepackage[utf8]{inputenc}
\usepackage[english]{babel}
\usepackage[babel]{csquotes}
\usepackage{amsmath}
\usepackage{textcomp}
\usepackage{gensymb}   % for \degree
\usepackage{graphicx}
\usepackage{authblk}   % Multiple authors with affiliations
\usepackage{sidecap}   % for captions next to figures
\sidecaptionvpos{figure}{c}

@@

\author[1,2,3,*]{\small Felix Molter}
\author[2,4]{\small Armin W. Thomas}
\author[5,6]{\small Scott A. Huettel}
\author[2,7]{\small Hauke R. Heekeren}
\author[1,2,3]{\small Peter N. C. Mohr}
\affil[1]{\footnotesize School of Business \& Economics, Freie Universität Berlin, Berlin, Germany}
\affil[2]{\footnotesize Center for Cognitive Neuroscience, Freie Universität Berlin, Berlin, Germany}
\affil[3]{\footnotesize WZB Berlin Social Science Center, Berlin, Germany}
\affil[4]{\footnotesize Department of Electrical Engineering and Computer Science, Technische Universität Berlin, Berlin, Germany}
\affil[5]{\footnotesize Center for Cognitive Neuroscience, Duke University, Durham, NC, USA}
\affil[6]{\footnotesize Department for Psychology and Neuroscience, Duke University, Durham, NC, USA}
\affil[7]{\footnotesize Department for Education and Psychology, Freie Universität Berlin, Berlin, Germany}
\affil[*]{\footnotesize E-mail: \href{mailto:felixmolter@gmail.com}{felixmolter@gmail.com}}

@@

% Fig. 1: Choice task and attribute space
\begin{SCfigure}[0.8]
\centering
\includegraphics[scale=1]{../figures/1-task_stimuli.pdf}
\caption{\textbf{Choice task and attribute space of stimuli.} \textbf{(a)} Choice task. Trials started with a central fixation cross for 1.5 s. Next, three gambles were presented. Participants made a choice by button press, without time limit. Eye movements were recorded during the choice phase. Finally, brief feedback on the choice (but not on the gamble outcome) was displayed. \textbf{(b)} Attribute space. Each gamble is described by two attributes: Probability $p$ and outcome $m$. The core options $A$ and $B$ were presented in every trial, along with one of four decoy alternatives: Compromise decoys ($C_A$ or $C_B$) and asymmetrically dominated decoys ($D_A$ or $D_B$) are expected to elicit compromise and attraction effects, respectively. Dashed lines indicate possible stimulus placements. \textbf{(c)} Exact positions for $C_B$, $A$ and $C_A$ were calibrated for pairwise indifference in separate binary choice estimation blocks for each individual. Differences in outcome between neighbouring alternatives was not less than 2 EUR. Dominated decoys $D_A$ and $D_B$ were always 2\% and 1 EUR worse than $A$ and $B$ respectively.}
\hspace{0.5cm}
\label{fig:task}
\end{SCfigure}

@@

The GLA variant also described 17 of 40 (42.5\%) participants best (Table \ref{tab:switchboard-best-ind}). Another similar variant with additive attribute integration described an additional four participants best. An additive relationship between attributes is typically assumed by models of multi-attribute multi-alternative choice \autocite{roe2001MultialternativeDecisionField,usher2004LossAversionInhibition}. Furthermore, additive integration of probability and outcome has recently been suggested as an alternative to multiplicative mechanisms and has been demonstrated to be equivalent for particular parameterisations and parameter values \autocite{rouault2019PrefrontalMechanismsCombining}. Five participants were best described by variants similar to the GLA, but using gaze-dependent leakage or inhibition instead of an alternative-wise gaze-discount. Note that gaze-dependent inhibition and leakage mechanisms can act similarly to an alternative-wise gaze-discount: All three mechanisms effectively reduce accumulated evidence of non-fixated alternatives. While the alternative-wise gaze bias mechanism applies a constant discount to the accumulation inputs, the gaze-dependent inhibition mechanism is proportional to the accumulated evidence of the currently fixated alternative, and applies to the already accumulated evidence of other options, not the inputs to the accumulation process. Gaze-dependent leakage similarly reduces already accumulated evidence, proportional to the momentary accumulator value.

@@

% Fig. 5: Hybrid variant
\begin{SCfigure}[0.6][tb]
\centering
\includegraphics[scale=1]{../figures/5-hybrid-variant.pdf}
\caption{\textbf{Hybrid model variant details.} \textbf{(a, b)} Number of participants better described by the hybrid variant (pink) or the GLA (grey), dependent on strength of attraction \textbf{(a)} and compromise \textbf{(b)} effects. Participants with strong attraction effects were better described by the hybrid variant. \textbf{(c, d)} Individual observed and predicted RST in attraction \textbf{(c)} and compromise trials \textbf{(d)}. Compared to GLA (Fig. \ref{fig:modelcomparison}i, j), the hybrid model better predicted strong attraction effects for some participants. Predictions of compromise effects are similar. \textbf{(e, f)} Observed and model-predicted probability of choosing the target alternative, depending on the target’s relative dwell time advantage. Like other gaze-dependent models (Fig. \ref{fig:modelcomparison}), the hybrid variant generally captured the positive association between gaze and choice. In contrast to GLA, however, it predicted an overall higher probability of choosing the target in attraction trials \textbf{(e)}. Predictions in compromise trials \textbf{(f)} are similar to GLA. White bars and error bars indicate mean $\pm$ s.e. observed data from even-numbered trials. Model predictions are based on 50 simulations for each odd-numbered trial.}
\label{fig:hybrid}
\end{SCfigure}

@@

\section*{Acknowledgements}
F.M. is supported by the International Max Planck Research School on the Life Course (LIFE). A.T. is supported by the Max Planck School of Cognition and Stanford Data Science.

\section*{Author contributions}
F.M., S.A.H. and P.N.C.M. designed the study. F.M. and A.W.T. conceptualized the behavioural models. F.M. performed the experiment, wrote all software, analysed the data and visualized results. P.N.C.M. and H.R.H. supervised the project. P.N.C.M. acquired funding. F.M. wrote the original draft. All authors reviewed and edited the manuscript.

\section*{Competing interests}
The authors declare no competing interests.

@@

% Supplementary Fig.: Dwell-regression weights
\begin{SCfigure}[0.8][t]
\centering
\includegraphics[scale=1]{../figures/S_dwell-regression_weights.pdf}
\caption{\textbf{Weight estimates from regression analyses on absolute dwell times.} We performed two mixed-effects regression analyses of dwell time towards each AOI onto stimulus properties: \textbf{(a, c)} Regressing the total dwell time towards an AOI over a full trial onto AOI column, row, attribute rank (best, middle or worst value on the attribute), two dummy predictors coding alternative, attribute (probability or outcome) and whether the AOI belonged to the subsequently chosen alternative. \textbf{(b, d)} Second, we binned dwell times in each trial into five time-bins and added an interaction term with time-bin for each predictor. The panels show the interaction term weights. Analyses were carried out separately for attraction \textbf{(a, b)} and compromise \textbf{(c, d)} trials. Regression models had random intercepts and slopes across participants. Points and intervals mark posterior mean estimates and associated HDI$_{95}$ (coloured green if the interval excluded 0).}
\label{fig:dwell-regression-weights}
\end{SCfigure}

@@

\begin{landscape}
\begin{table}[ht]

@@

\end{table}
\end{landscape}
